\MakeUppercase{Layered $H$-Partitions of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-05-09 k-planar.tex
Hmmmm

Let $G$ be a $k$-plane graph and let $G^+$ be the graph obtained as follows:
\begin{compactenum}
  \item Make $G$ planar by adding dummy vertices at each crossing.
  \item Add a $K_4$ that contains $G$ in one of its inner faces $F_0$.
  \item Add edges to complete the resulting graph to a triangulation.
\end{compactenum}
Let $r$ be the vertex of the $K_4$ not incident on $F_0$ and let $T$ be a BFS spanning tree of $G^+$ rooted at $r$.    Observe that, for any edge $vw\in E(G)$, $\dist_{G^+}(v,w) \le k+1$ and therefore, for any two vertices $v,w\in V(G)$,
\begin{equation} 
  \dist_{G^+}(v,w) \le (k+1)\cdot \dist_G(v,w) \eqlabel{distance-preserving}
\end{equation}
For each integer $i\ge 0$, let $V_i=\{v\in V(G^+): \dist_{G^+}(r,v)=i\}$ and observe that $V_0,V_1,\ldots,V_{t}$ is a layering of $G^+$ (but not of $G$).

Let $K=K(G^+,T,r)$ be a PS-tree and observe that the root $r'$ of $K$ has $F_{r'}=F_0$.  This implies:
\begin{compactenum}[(PR6)]
  \item \proplabel{all-in}  $V(G)\subseteq V(N_{r'}-F_{r'})$.
\end{compactenum}

For any pair of vertices $v,w\in V(G)$, consider the set $A_{vw}=\{a\in V(K): v,w\in V(N_a-F_a) \}$.  By \propref{all-in}, $A_{vw}$ contains the root $r'$ of $K$.  By \propref{children}, any node $a\in A_{vw}$ has at most one child in $A_{vw}$ and any node $a\not\in A_{vw}$ has no child in $A_{vw}$.  It follows that $K[A_{vw}]$ is a vertical path in $K$ that contains the root.  We define $a(vw)$ to be the lower endpoint of $K[A_{vw}]$.

For any vertex $v\in V(G)$ define $A_v=\bigcap\{vw\in E(G)} A_{vw}$.  $K[A_v]$ is a vertical path in $K$ that contains the root.  We define $a(v)$ to be the lower endpoint of $K[A_v]$.

For every node $x\in V(K)$, we define the \emph{separator} of $X$ as
\[
   S_x = \{v\in V(G): a(v)=x \} \enspace .
\]
Clearly $\mathcal{S}=\{S_x:x\in V(K)\}$ is a partition of $V(G)$.

@@

Next we describe a tree decomposition $(B_x:x\in V(K))$ of $H$.  We deliberately use the same node set $V(H)=V(K)$ here, because the subtree $K[x]:=K[\{z:x\in B_z\}]$ will be rooted at $x$.  We claim that, for every edge $xy\in V(H)$, one of $x$ or $y$ is a $K$-ancestor of the other.
Indeed, since $xy\in E(H)$, there exists an edge $vw\in E(G)$ such that $v\in S_x$ and $w\in S_y$.  By definition, $x\in A_v\subseteq A_{vw}$ and $y\in A_w\subseteq A_{vw}$.  Since $K[A_{vw}]$ is a vertical path in $K$, it follows that one of $x$ or $y$ is a $K$-ancestor of the other.

To describe our tree decomposition we define, for each $y\in V(H)$, the subtree $K[y]=K[\{x:y\in B_x\}]$.  The subtree $K[y]$ contains $y$.  Additionally, for every edge $xy\in E(H)$, where $y$ is an ancestor of $x$, $K[y]$ includes the path from $y$ to $x$ in $K$.  That this produces a tree-decomposition is due to the fact, established in the preceding paragraph, that for every edge $xy$ of $H$, $x$ is a $K$-ancestor of $y$ (or vice-versa).  What remains is to upper bound the width of this decomposition.

Now, fix some node $x\in V(K)$ and consider the contents of $B_x$.  As we have just argued, $B_x$ contains only $K$-ancestors of $x$ (including $x$ itself). If $B_x$ contains some $K$-ancestor $y$ of $x$, it is because there is an edge $vw\in E(G)$ with $v\in S_x$ and $w\in S_y$. Since $v\in S_x$, $v\in V(N_x-F_x)$. Since, $w\in S_y$, $w\in V(N_y-F_y)$.

Now, consider the graph $A$ that contains $F_{a}$ for all $K$-ancestors $a$ of $x$. See \figref{A}.  It is helpful to think of this graph as follows:  We begin with $F_{a_0}$ where $a_0=r'$ is the root of $K$.  We then cut $F_{a_0}$ into two pieces $F^{\bar{x}}_{a_0}$ and $F^x_{a_0}$ where $F^x_{a_0}$ is the piece that contains $F_x$. (The path that makes this cut contains two legs $Q_i$ and $Q_j$ of a tripod $Y$ and one edge $v_iv_j$ of its central triangle $\tau$.)  This process continues by partitioning $F_{a_1}:=F^x_{a_0}$ recursively.  The process terminates after some number $r$ of iterations, when $F^x_{a_r}=F_x$.

@@

The edge $vw$ corresponds to a walk $v=v_0,v_1,\ldots,v_{\ell}$ in $G^+$ that,
(i)~begins at $v\in V(N_x-F_x)$, (ii)~contains only vertices in $V(N_y-F_y)$, (iii)~ends at $w\in V(N_y-F_y)$, (iv)~has length $\ell\le k+1$.
\begin{compactenum}[(i)]
  \item $v\in V(N_x)$ since $x\in A_v$ and $v\in V(N_a-F_a)$ for all $a\in A_v$;

  \item $v_0,v_1,\ldots,v_{\ell}\in V(N_a-F_a)$ since, otherwise $v_i\in V(F_y)$.  By \propref{ancestor-boundary}, $v_i\in Y_{a}$ for some $K$-ancestor $a\neq y$ of $y$.

  The curve $vz'$ remains in the interior of $F_a$ since, if it intersected the boundary of $F_a$, then one of $vw$ or $wz$ would intersect $Y_{a'}$ for some ancestor $a'\neq a$ of $a$.  If $vw$ contains a vertex of $Y_{a'}$, then $v\not\in S_x$.  If $wz$ contains a vertex of $Y_{a'}$ then $w,z\not\in S_a$.

  \item $w\in V(N_y)$ since $y\in A_w$ and $w\in V(N_a-F_a)$ for all $a\in A_w$;

  \item That $vz'$ eventually reaches $F^{\bar{x}}_a$ follows from the fact that, in either case (a) or (b), $vz$ contains a vertex of $Y_a$.  The first such vertex encountered walk walking along $vz$ is the endpoint $z'$.

  \item That $vz'$ contains at most $2k+1$ vertices of $A$ follows from the fact that $vz'$ is contained in a path of length at most 2 in $G$, and $G$ is $k$-planar.
\end{compactenum}
We call any curve with properties (i)--(iv) an \emph{admissible $xa$-curve} (for $A$).  Therefore, to upper-bound $|B_x|$, it suffices to upper-bound the number of faces $F^{\bar{x}}_a$ of $A$ such that there exists an admissible $xa$-curve. To avoid triple-subscripts, we will say that an $xa$-curve \emph{visits} $y\in N(K)$ if it contains a vertex of $F^{\bar{x}}_y$.  It is furthermore sufficient to consider $xa$-curves that are \emph{minimal} in the sense that, if the curve visits $x=x_0,x_1,\ldots,x_\ell=a$ in this order, then $x_i\neq a$ for any $i\in\{0,\ldots,\ell-1\}$.

@@

Consider some minimal admissible $xa$-curve that visits $x=x_0,x_1,\ldots,x_\ell=a$ in this order.  Refer to \figref{colouring}.  Let $z_0,\ldots,z_{\ell'}$ be the subsequence of $x_0,\ldots,x_\ell$ consisting only of those $x_i$ that are strict $K$-ancestors of each of $x_0,\ldots,x_{i-1}$.  This yields a colour sequence $c_1,\ldots,c_{\ell'}$, $\ell'\le \ell$, where $c_i=c(z_{i-1},z_i)$ for each $i\in\{1,\ldots,\ell'\}$.   There are at most $3^{\ell'}$ such colour sequences and, if an admissible $xa$-curve has the same colour sequence as a minimal admissible $xa'$-curve then $a=a'$.  Therefore, for some fixed $x\in V(K)$, the $K$-ancestors $a$ of $x$ such that there exists a minimal admissible $xa$-curve is at most
\begin{equation}
   |B_x| \le \sum_{\ell=0}^{2k+1} 3^\ell = (3^{2k+2}-1)/2 
       = (9^{k+1}-1)/2 \enspace .  \eqlabel{bag-size}
\end{equation}

This is all we need to complete the proof of:

>>>>

Let $G$ be a $k$-plane graph and let $G^+$ be the graph obtained as follows:
\begin{compactenum}
  \item Make $G$ planar by adding dummy vertices at each crossing to obtain a graph $G_1$.
  \item Add a $K_4$ that contains $G_1$ in one of its inner faces $F_0$ to obtaina graph $G_2$.
  \item Add edges to complete $G_2$ to a triangulation $G^+$.
\end{compactenum}
Let $r$ be the vertex of the $K_4$ not incident on $F_0$ and let $T$ be a BFS spanning tree of $G^+$ rooted at $r$.    Observe that, for any edge $vw\in E(G)$, $\dist_{G^+}(v,w) \le k+1$ and therefore, for any two vertices $v,w\in V(G)$,
\begin{equation} 
  \dist_{G^+}(v,w) \le (k+1)\cdot \dist_G(v,w) \eqlabel{distance-preserving}
\end{equation}
For each integer $i\ge 0$, let $V_i=\{v\in V(G^+): \dist_{G^+}(r,v)=i\}$ and observe that $V_0,V_1,\ldots,V_{t}$ is a layering of $G^+$ (but not of $G$).

Let $K=K(G^+,T,r)$ be a PS-tree and observe that the root $r'$ of $K$ has $F_{r'}=F_0$.  This implies:
\begin{compactenum}[(PR6)]
  \item \proplabel{all-in}  $V(G_1)\subseteq V(N_{r'}-F_{r'})$.
\end{compactenum}

For each vertex $v\in V(G_1)$, let $P_v=\{a\in V(K): v\in V(N_a-F_a)\}$.  
By (PR6), $P_{v}$ contains the root $r'$ of $K$.  By \propref{children}, any node $a\in P_{v}$ has at most one child in $P_{v}$ and any node $a\not\in P_{v}$ has no child in $P_{v}$.  It follows that $K[P_{v}]$ is a vertical path in $K$ that contains the root.

For each edge $vw\in E(G)$, consider the walk $W_{vw}$ in $G^+$ that corresponds to $vw$ and define $A_{vw}=\bigcap_{z\in W_{vw}} P_z$.  Since $W_{vw}\subseteq V(G_1)$, $A_{vw}$ is well-defined.  For each $z\in W_{vw}$, $P_{z}$ contains $r'$ and is a vertical path in $K$.  The common intersection of two or more vertical paths is a vertical path.  Therefore $A_{vw}$ is a non-empty vertical path in $K$.  For any vertex $v\in V(G)$ define $A_v=\bigcap_{vw\in E(G)} A_{vw}$.  Again, $K[A_v]$ is a vertical path in $K$ that contains $r'$.  We define $a(v)$ to be the lower endpoint of $K[A_v]$.

For every node $x\in V(K)$, define the \emph{separator}
\[
   S_x = \{v\in V(G): a(v)=x \} \enspace .
\]
Clearly $\mathcal{S}=\{S_x:x\in V(K)\}$ is a partition of $V(G)$.

@@

Next we describe a tree decomposition $(B_x:x\in V(K))$ of $H$.  We deliberately use the same node set $V(H)=V(K)$ here, because the subtree $K[x]:=K[\{z:x\in B_z\}]$ will be rooted at $x$.

\begin{clm}\clmlabel{xy-ancestor}
   For every edge $xy\in V(H)$, one of $x$ or $y$ is a $K$-ancestor of the other.
 \end{clm}

 \begin{proof}
   Since $xy\in E(H)$, there exists an edge $vw\in E(G)$ such that $v\in S_x$ and $w\in S_y$.  By definition, $x\in P_v\subseteq A_{vw}$ and $y\in P_w\subseteq A_{vw}$.  Since $K[A_{vw}]$ is a vertical path in $K$, it follows that one of $x$ or $y$ is a $K$-ancestor of the other. 
\end{proof}

The tree decomposition $(B_x:x\in V(K))$ of $H$ is defined as follows: For each edge $xy$ where $y$ is a $K$-ancestor of $x$, we add $y$ to be $B_{x'}$ for every node $x'$ on the path from $x$ to $y$ in $K$.  That this produces a tree-decomposition is due to \clmref{xy-ancestor}.  What remains is to upper bound the width of this decomposition.

\begin{clm}
  The tree decomposition $\{B_x: x\in V(K)\}$ of $H$ contains no bag larger than BLAH.
\end{clm}

\begin{proof}
  Fix some node $x\in V(K)$ and consider the contents of $B_x$.  By \clmref{xy-ancestor}, $B_x$ contains only $K$-ancestors of $x$ (including $x$ itself). If $B_x$ contains some $K$-ancestor $y$ of $x$, it is because there is an edge $vw\in E(G)$ with $v\in S_x$ and $w\in S_y$.

  Consider the graph $A$ that contains $F_{a}$ for all $K$-ancestors $a$ of $x$. See \figref{A}.  It is helpful to think of this graph as follows:  We begin with $F_{a_0}$ where $a_0=r'$ is the root of $K$.  We then cut $F_{a_0}$ into two pieces $F^{\bar{x}}_{a_0}$ and $F^x_{a_0}$ where $F^x_{a_0}$ is the piece that contains $F_x$. (The path that makes this cut contains two legs $Q_i$ and $Q_j$ of a tripod $Y$ and one edge $v_iv_j$ of its central triangle $\tau$.)  This process continues by partitioning $F_{a_1}:=F^x_{a_0}$ recursively.  The process terminates after some number $r$ of iterations, when $F^x_{a_r}=F_x$.

@@

  The edge $vw$ corresponds to a walk $W_{vw}=v_0,v_1,\ldots,v_{\ell}$ in $G^+$ that
  (i)~begins at $v_0=v\in V(N_x-F_x)$, (ii)~contains only vertices in $V(N_y-F_y)$, (iii)~ends at $w\in V(N_y)$, \note{PM:notation!} (iv)~has length $\ell\le k+1$.
  \begin{compactenum}[(i)]
    \item $v\in V(N_x-F_x)$ since $x\in P_v$ and $v\in V(N_a-F_a)$ for all $a\in P_v$;

    \item $v_0,v_1,\ldots,v_{\ell}\in V(N_y-F_y)$ since, otherwise there exists some $i\in\{1,\ldots,v_k\}$ such that $v_i\not\in V(N_y-F_y)$.  But this implies that $y\not\in P_{v_i}$ which implies that $y\not\in A_{vw}$ which implies that $y\not\in A_w$, a contradiction.

    \item That $V_{wv}$ contains a vertex $w'\in F^\bar{x}_{y}$ is due to the fact that $y\in P_w$

    and $w\in V(N_y)$.

    That $vz'$ eventually reaches $F^{\bar{x}}_a$ follows from the fact that, in either case (a) or (b), $vz$ contains a vertex of $Y_a$.  The first such vertex encountered walk walking along $vz$ is the endpoint $z'$.

    \item That $\ell\le k+1$ follows from the fact that $G$ is $k$-planar and that each of $v_1,\ldots,v_{\ell-1}$ corresponds to a crossing involving $vw$.  Therefore $\ell-1\le k$.
  \end{compactenum}
  We call any curve with properties (i)--(iv) an \emph{admissible $xa$-curve} (for $A$).  Therefore, to upper-bound $|B_x|$, it suffices to upper-bound the number of faces $F^{\bar{x}}_a$ of $A$ such that there exists an admissible $xa$-curve. To avoid triple-subscripts, we will say that an $xa$-curve \emph{visits} $y\in N(K)$ if it contains a vertex of $F^{\bar{x}}_y$.  It is furthermore sufficient to consider $xa$-curves that are \emph{minimal} in the sense that, if the curve visits $x=x_0,x_1,\ldots,x_\ell=a$ in this order, then $x_i\neq a$ for any $i\in\{0,\ldots,\ell-1\}$.

@@

  Consider some minimal admissible $xa$-curve that visits $x=x_0,x_1,\ldots,x_\ell=a$ in this order.  Refer to \figref{colouring}.  Let $z_0,\ldots,z_{\ell'}$ be the subsequence of $x_0,\ldots,x_\ell$ consisting only of those $x_i$ that are strict $K$-ancestors of each of $x_0,\ldots,x_{i-1}$.  This yields a colour sequence $c_1,\ldots,c_{\ell'}$, $\ell'\le \ell$, where $c_i=c(z_{i-1},z_i)$ for each $i\in\{1,\ldots,\ell'\}$.   There are at most $3^{\ell'}$ such colour sequences and, if an admissible $xa$-curve has the same colour sequence as a minimal admissible $xa'$-curve then $a=a'$.  Therefore, for some fixed $x\in V(K)$, the $K$-ancestors $a$ of $x$ such that there exists a minimal admissible $xa$-curve is at most
  \begin{equation}
     |B_x| \le \sum_{\ell=0}^{2k+1} 3^\ell = (3^{2k+2}-1)/2 
         = (9^{k+1}-1)/2 \enspace .  \eqlabel{bag-size}
  \end{equation}
\end{proof}
This is all we need to complete the proof of:
